How not to constructivize cohomology

### 2018-10-16 slides-pssl104.tex
Final version of the PSSL slides with notes

      \footnotesize
      \mbox{``Every map~$\NN \to \NN$ is computable.''}
      \mbox{``Every map~$\RR \to \RR$ is continuous.''}
      \mbox{``Every map~$\affl \to \affl$ is polynomial.''}
      \mbox{``Heyting Arithmetic has exactly one model.''}
      \mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}
      \mbox{``There is an injection~$\RR \to \NN$.''} \\
      \rotatebox{90}{\dots}
    \end{column}

@@

      \footnotesize
      \mbox{program extraction} \\
      \mbox{synthetic differential geometry} \\
      \mbox{synthetic algebraic geometry} \\
      % \mbox{synthetic probability theory} \\
      \mbox{synthetic domain theory} \\
      \mbox{new reduction techniques in algebra} \\
      \mbox{Bohr topos for quantum mechanics} \\
      \rotatebox{90}{\dots}
    \end{column}
  \end{columns}

@@

  \jnote{2-}{The internal language of a topos~$\E$ is a device which defines
  for any formula~$\varphi$ of a certain language (a form of extensional type
  theory) what it means for~$\varphi$ to hold \emph{internally to~$\E$},
  written~``$\E \models \varphi$''. This translation process is sound with
  respect to intuitionistic logic; hence any theorem of constructive
  mathematics is valid in any topos. Only few toposes validate classical logic
  (for instance~$\Sh(X)$ does if~$X$ is a discrete space and the law of
  excluded middle is available in the metatheory).

@@

  % XXX: Cite Shulman and ABSS. Also Streicher's lecture notes.

  \jnote{3-}{The intermediate value theorem (``any continuous function with
  opposite signs has a zero'') doesn't admit a constructive proof, because for
  most spaces~$X$ the external translation~$\Sh(X) \models \text{IVT}$ is not
  true -- it's not true that in continuous families of continuous functions
  with opposite signs, zeros can locally be picked continuously, as
  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

  Over reduced schemes, every finite type sheaf of modules is finite locally free on a
  dense open. This statement (``important hard exercise'' 13.7.K in
  \fixedhref{http://math.stanford.edu/~vakil/216blog/FOAGnov1817public.pdf}{[Vakil
  2017]}) is just the external translation of the easy-to-prove internal
  statement that every finitely generated vector space does \emph{not not}
  admit a finite basis. (A scheme is reduced if and only if its structure sheaf
  looks like a field from the internal point of view (in the sense that $1 \neq
  0$ and~$\neg(\text{$x$ invertible}) \Rightarrow x = 0$). This is why the
  reducedness condition is important.)}
\end{frame}

@@

  A basic application is the following. Any student in algebraic geometry
  needs, at some point in her life, to compute the cohomology of projective space.
  At a later point she needs to compute higher direct images
  along~$\mathbb{P}^n_S \to S$, where~$\mathbb{P}^n_S$ is a relative version of
  projective space. Since higher direct images are just internal
  sheaf cohomology, she can in fact skip the second computation.}
\end{frame}

@@

  In particular, we hope to prove the conjecture that the statement ``any
  module embeds into a flabby module'' is not constructively provable by
  verifying that it doesn't hold in the effective topos.

@@

  Details on the self-referential construction~``$K \defeq \{ f(a_K) \}$'' are in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-flabby-objects.pdf}{this
  draft paper}.

  A module is~$\neg\neg$-separated if and only if~$\neg\neg(x = 0)$ implies~$x = 0$.}
\end{frame}

% XXX: ctcl.pdf, realizability book

@@

  Hence it appears that the simple idea of basing sheaf cohomology on flabby
  resolutions doesn't work, constructively. More work is needed to give a
  constructive account of sheaf cohomology and, more generally, derived functors.

  I hope that some day, we can study the cohomology of the the smallest dense
  sublocale of the one-point space.$^\star$

  \bigskip
  \bigskip
  \bigskip
  \bigskip

@@

  An alternative way to phrase this question is as follows.
  Let~$\Set_{\neg\neg}$ be the smallest dense subtopos of~$\Set$, the topos of
  double-negation sheaves. The forgetful functor~$\mathrm{Ab}(\Set_{\neg\neg})
  \to \mathrm{Ab}$ is left-exact, but might not be right-exact, since
  a map~$f : A \to B$ is an epimorphism
  in~$\mathrm{Ab}(\Set_{\neg\neg})$ if and only if~$\forall y \in B\_
  \neg\neg(\exists x \in A\_ f(x) = y)$, which is weaker than being surjective.
  What do its right derived functors look like?\par}
\end{frame}
\addtocounter{framenumber}{-1}

>>>>

      \footnotesize
      \mbox{``Every map~$\NN \to \NN$ is computable.''}
      \mbox{``Every map~$\RR \to \RR$ is continuous.''}
      \mbox{``Every map~$\affl \to \affl$ is polynomial.''}
      \href{https://www.staff.science.uu.nl/~ooste110/realizability/arithcatsubmit.pdf}{\mbox{``Heyting Arithmetic has exactly one model.''}}
      \href{http://www1.maths.leeds.ac.uk/~pmtng/Slides/gambino-lisbon-jan2018.pdf}{\mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}}
      \href{https://rawgit.com/iblech/mathezirkel-kurs/master/superturingmaschinen/slides-warwick2017.pdf}{\mbox{``There is an injection~$\RR \to \NN$.''}} \\
      \rotatebox{90}{\dots}
    \end{column}

@@

      \footnotesize
      \mbox{program extraction} \\
      \href{http://math.andrej.com/2008/08/13/intuitionistic-mathematics-for-physics/}{\mbox{synthetic differential geometry}} \\
      \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\mbox{synthetic algebraic geometry}} \\
      \href{https://www.dpmms.cam.ac.uk/~martin/Research/Oldpapers/synthetic91.pdf}{\mbox{synthetic domain theory}} \\
      \href{https://rawgit.com/iblech/internal-methods/master/slides-bayreuth2018.pdf}{\mbox{new reduction techniques in algebra}} \\
      \href{https://rawgit.com/iblech/internal-methods/master/}{\mbox{Bohr topos for quantum mechanics}} \\
      \rotatebox{90}{\dots}
    \end{column}
  \end{columns}

@@

  \jnote{2-3}{The internal language of a topos~$\E$ is a device which defines
  for any formula~$\varphi$ of a certain language (a form of extensional type
  theory) what it means for~$\varphi$ to hold \emph{internally to~$\E$},
  written~``$\E \models \varphi$''. This translation process is sound with
  respect to intuitionistic logic; hence any theorem of constructive
  mathematics is valid in any topos. Only few toposes validate classical logic
  (for instance~$\Sh(X)$ does if~$X$ is a discrete space and the law of
  excluded middle is available in the metatheory).

@@

  \jnote{3}{The intermediate value theorem (``any continuous function with
  opposite signs has a zero'') doesn't admit a constructive proof, because for
  most spaces~$X$ the external translation~$\Sh(X) \models \text{IVT}$ is not
  true -- it's not true that in continuous families of continuous functions
  with opposite signs, zeros can locally be picked continuously, as
  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

  Over reduced schemes, every finite type sheaf of modules is finite locally free on a
  dense open. This statement (``important hard exercise'' 13.7.K in
  \fixedhref{http://math.stanford.edu/~vakil/216blog/FOAGnov1817public.pdf}{[Vakil
  2017]}) is just the external translation of the easy-to-prove internal
  statement that every finitely generated vector space does \emph{not not}
  admit a finite basis. (A scheme is reduced if and only if its structure sheaf
  looks like a field from the internal point of view (in the sense that $1 \neq
  0$ and~$\neg(\text{$x$ invertible}) \Rightarrow x = 0$). This is why the
  reducedness condition is important.)}

  \jnote{4-}{Excellent references on the internal language include:
  \begin{center}
    \fixedhref{https://www.oliviacaramello.com/Papers/Papers.htm}{\includegraphics[height=3.5cm]{olivia-tst}}
    \fixedhref{https://store.doverpublications.com/0486450260.html}{\includegraphics[height=3.5cm]{goldblatt-logic}}
    \fixedhref{https://www.worldcat.org/title/sheaves-in-geometry-and-logic-a-first-introduction-to-topos-theory/oclc/24428855}{\includegraphics[height=3.5cm]{moerdijk-mac-lane-sheaves}}
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{\includegraphics[height=3.5cm]{streicher-ctcl}}
  \end{center}

  We use an extension of the original form of the internal language which
  allows for unbounded quantification, Mike Shulman's
  \fixedhref{https://arxiv.org/abs/1004.3802}{\emph{stack semantics}}.
  (Independently, Steve Awodey, Carsten Butz, Alex Simpson and Thomas Streicher
  developed a
  \fixedhref{http://homepages.inf.ed.ac.uk/als/Research/Sources/set-models.pdf}{similar
  semantics}.)}
\end{frame}

@@

  A basic application is the following. Any student in algebraic geometry
  needs, at some point in her life, to compute the cohomology of projective space.
  At a later point she needs to compute higher direct images
  along~$\mathbb{P}^n_S \to S$, where~$\mathbb{P}^n_S$ is a relative version of
  projective space. Since higher direct images are just internal
  sheaf cohomology, she can in fact skip the second computation.

  Further progress along these lines is hindered by the fact that we don't yet
  have a constructive account of sheaf cohomology.}
\end{frame}

@@

  In particular, we hope to prove the conjecture that the statement ``any
  module embeds into a flabby module'' is not constructively provable by
  verifying that it doesn't hold in the ef{}fective topos.

@@

  Details on the self-referential construction~``$K \defeq \{ f(a_K) \}$'' are in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-flabby-objects.pdf}{this
  draft paper}. A module is~$\neg\neg$-separated if and only if~$\neg\neg(x = 0)$ implies~$x = 0$.
  References on the ef{}fective topos include Martin Hyland's
  \fixedhref{https://webdpmms.maths.cam.ac.uk/~martin/Research/Oldpapers/hyland-effectivetopos.pdf}{survey
  paper} and the canonical book by one of our honoraries:

  {\centering\fixedhref{https://www.staff.science.uu.nl/~ooste110/boekbegin.pdf}{\includegraphics[height=3.5cm]{oosten-realizability}}\par}}
\end{frame}

@@

  Hence it appears that the simple idea of basing a constructive account of sheaf cohomology on flabby
  resolutions doesn't work.

  More work is needed.
  I hope that some day, we can study the cohomology of the smallest dense
  sublocale of the one-point space.$^\star$

  {\centering\includegraphics[height=2cm]{phantoms}\par}

@@

  An alternative way of putting this question is as follows.
  Let~$\Set_{\neg\neg}$ be the smallest dense subtopos of~$\Set$, the topos of
  double negation sheaves. The forgetful functor~$\mathrm{Ab}(\Set_{\neg\neg})
  \to \mathrm{Ab}$ is left-exact, but might not be right-exact, since
  a map~$f : A \to B$ is an epimorphism
  in~$\mathrm{Ab}(\Set_{\neg\neg})$ if and only if~$\forall y \in B\_
  \neg\neg(\exists x \in A\_ f(x) = y)$, which is weaker than being surjective.
  What do its right derived functors look like?\par}
\end{frame}
\addtocounter{framenumber}{-1}
